Synthesising Recursive Functions for First-Order Model Counting:\\ Challenges, Progress, and Conjectures

### 2023-03-03 doc/paper/paper.tex
Describe one more instance

% TODO: (everywhere in the paper): emphasise that the experimental results show
% Crane to be strictly more capable than ForcLift in a sense that it can solve
% everything that ForcLift can and more (including instances beyond C^2).

@@

% However, none of the publicly available exact (W)FOMC algorithms can efficiently
% compute functions as simple as a factorial.\footnote{There is some previous work
%   discussing FOMC for
%   factorials~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/kr/BremenK21,DBLP:conf/ijcai/Broeck16}.
%   One of them~\cite{DBLP:conf/ijcai/Broeck16} is an example `by hand' that does
%   not directly correspond to any algorithm.
%   Another~\cite{DBLP:journals/jair/Kuzelka21} proves that the problem is
%   liftable but contains no experimental results. The
%   third~\cite{DBLP:conf/kr/BremenK21} contains experimental results performed
%   with an algorithm that is not publicly available.}
% Two more papers to mention:
% 1. The L2C paper has the factorial example as well. But then why doesn't L2C
% work on it?
% 2. \cite{DBLP:journals/corr/abs-2302-04606} computes n! using publicly
% unavailable code

@@

\section{Experimental Results}\label{sec:results}
% TODO: rename 'experimental results' -> 'complexity results' (not just the
% title of the section but review every mention of the word 'experimental')
% TODO: make it clear that I'm not doing this by hand anymore, but I'm still
% removing multiplications by one.
% TODO (here or somewhere else): emphasise that there are two parts: search
% (with some artificial limit) + inference (whose complexity we can easily
% determine)
% TODO: describe ternary.mln as an example instance that's not in C^2 that can
% be lifted
% TODO: say that 'f' will always denote the main function

\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & Best Known & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!25}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{2}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \rowcolor{gray!25}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!25}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!25}\cmark & \xmark & & ${\min\{\, m, n \,\}}^2$ & --- & $mn$ \\
    \rowcolor{gray!25}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & Bijections & $m$ & --- & $m$ \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions,
    where $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that the algorithm was not able to find a solution.}\label{tbl:results}
\end{table*}
% TODO: 'Asymptotic Complexity of Counting' -> 'Complexity of \mathcal{F}_\phi'
% (introduce \mathcal{F}_\phi in the introduction, together with the flow graph)

@@

% TODO: emphasise that the reader should focus on the last two columns
% TODO: the caption sentence about not finding a solution could be confusing
% TODO: 'best known' sounds disappointing. Find a better alternative.

@@

% \footnote{The search tree has a high
%   branching factor, so exploring all nodes at depth five takes at most a few
%   seconds, whereas doing the same for depth six can be computationally
%   infeasible in some cases.}

@@

The experimental results are summarised in \cref{tbl:results}, where we compare
the solutions found by both algorithms to the best known ways of computing the
same quantities. The best-known asymptotic complexity for computing total
surjections is by Earnest~\shortcite{30049}. All other best-known complexity
results are inferred from the formulas and programs on the on-line encyclopedia
of integer sequences~\cite{oeis}. On instances that could already be solved by
\textsc{ForcLift}, the two algorithms perform equally well. However,
\textsc{Crane} can also solve all but one of the instances that
\textsc{ForcLift} fails on in at most cubic time. See supplementary material for
the exact solutions produced by \textsc{Crane} that correspond to the complexity
results in \cref{tbl:results}. We also note that standard (W)FOMC benchmarks
such as \cref{example:smokers} that are already supported by \textsc{ForcLift}
remain feasible for \textsc{Crane} as well. Moreover, our algorithm can also
handle the linear order axiom, for which a bespoke algorithm was recently
proposed by T{\'{o}}th and
Ku\v{z}elka~\shortcite{DBLP:journals/corr/abs-2211-01164}.

@@

In this paper, we showed how a state-of-the-art (W)FOMC algorithm can be
empowered by generalising domain recursion and adding support for cycles in the
graph that encodes a solution. To construct such graphs, \textsc{Crane}
supplements \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} with three
new compilation rules. Our experiments revealed a range of counting problems
that are liftable to \textsc{Crane} but not to any other (W)FOMC algorithm.
Although in this paper we focus on unweighted counting, \textsc{ForcLift}'s
support for weights trivially transfers to \textsc{Crane} as well. The common
thread across these newly liftable problems is (partial) injectivity. Thus, we
can formulate the following conjecture.

@@

>>>>

% TODO: (everywhere in the paper): emphasise that the (complexity) results show
% Crane to be strictly more capable than ForcLift in a sense that it can solve
% everything that ForcLift can and more (including instances beyond C^2).

@@

@@

\section{Complexity Results}\label{sec:results}
% TODO: make it clear that I'm not doing this by hand anymore, but I'm still
% removing multiplications by one.

% TODO: say that 'f' will always denote the main function

% TODO (here or somewhere else): emphasise that there are two parts: search
% (with some artificial limit) + inference (whose complexity we can easily
% determine)

% TODO: I think I need to mention what m and n are in the text, not just in the caption.

\begin{table*}[t]
  \centering
  \begin{tabular}{cccccc}
    \toprule
    \multicolumn{3}{c}{Function Class} & \multicolumn{3}{c}{Asymptotic Complexity of Counting} \\
    Partial & Endo- & Class & By Hand & With \textsc{ForcLift} & With \textsc{Crane} \\
    \midrule
    \rowcolor{gray!25}\cmark/\xmark & \cmark/\xmark & Functions & $\log m$ & $m$ & $m$ \\
    \xmark & \xmark & \multirow{2}{*}{Surjections} & $n \log m$ & $m^{3}+n^{3}$ & $m^{3}+n^{3}$ \\
    \xmark & \cmark & & $m \log m$ & $m^{3}$ & $m^{3}$ \\
    \rowcolor{gray!25}\xmark & \xmark & & $m$ & --- & $mn$ \\
    \rowcolor{gray!25}\xmark & \cmark & & $m$ & --- & $m^3$ \\
    \rowcolor{gray!25}\cmark & \xmark & & ${\min\{\, m, n \,\}}^2$ & --- & $mn$ \\
    \rowcolor{gray!25}\cmark & \cmark & \multirow{-4}{*}{Injections} & $m^2$ & --- & --- \\
    \xmark & \xmark & Bijections & $m$ & --- & $m$ \\
    \bottomrule
  \end{tabular}
  \caption{The worst-case complexity of counting various types of functions,
    where $m$ is the size of domain $\Gamma$, and $n$ is the size of domain
    $\Delta$. All asymptotic complexities are in $\Theta(\cdot)$. A dash means
    that the algorithm was not able to find a solution. In the case of
    \textsc{ForcLift}, this means that the greedy search algorithm ended with a
    formula not suitable for any compilation rule. In the case of
    \textsc{Crane}, this means that a complete solution could not be found after
    having explored the maximum allowed depth of the search
    tree.}\label{tbl:results}
\end{table*}
% TODO: 'Asymptotic Complexity of Counting' -> 'Complexity of \mathcal{F}_\phi'
% (introduce \mathcal{F}_\phi in the introduction, together with the flow graph)

@@

% TODO: emphasise that the reader should focus on the last two columns

@@

@@

The results are summarised in \cref{tbl:results}, where we compare the solutions
found by both algorithms to the manually-constructed ways of computing the same
functions that have been proposed before. In particular, the complexity for
computing total surjections is by Earnest~\shortcite{30049}, and all other such
complexities are inferred from the formulas and programs on the on-line
encyclopedia of integer sequences~\cite{oeis}. On instances that could already
be solved by \textsc{ForcLift}, the two algorithms perform equally well.
However, \textsc{Crane} can also solve all but one of the instances that
\textsc{ForcLift} fails on in at most cubic time. See supplementary material for
the exact solutions produced by \textsc{Crane} that correspond to the complexity
results in \cref{tbl:results}.

\paragraph{Miscellaneous other instances.}
We also note that standard (W)FOMC benchmarks such as \cref{example:smokers}
that are already supported by \textsc{ForcLift} remain feasible for
\textsc{Crane} as well. Moreover, our algorithm can also handle the linear order
axiom, for which a bespoke algorithm was recently proposed by T{\'{o}}th and
Ku\v{z}elka~\shortcite{DBLP:journals/corr/abs-2211-01164}. To demonstrate that
\textsc{Crane} is capable of finding efficient solutions to problems beyond
known domain-liftable fragments such as \Ctwo{}, we run \textsc{Crane} in greedy
search mode on the following formula:
\begin{equation}\label[formula]{eq:ternary}
  \begin{split}
    &\begin{multlined}[t]
      (\forall X,W \in \Gamma\text{. }\forall Y \in \Delta\text{. }\forall Z \in \Lambda\text{. }\\
      \texttt{p}(X, Y, Z) \land \texttt{p}(W, Y, Z) \Rightarrow X = W) \land{}
    \end{multlined}\\
    &\begin{multlined}[t]
      (\forall X \in \Gamma\text{. }\forall Y,W \in \Delta\text{. }\forall Z \in \Lambda\text{. }\\
      \texttt{p}(X, Y, Z) \land \texttt{p}(X, W, Z) \Rightarrow Y = W).
    \end{multlined}
  \end{split}
\end{equation}
For every element $z$ of $\Lambda$, \cref{eq:ternary} restricts
$\texttt{p}(X, Y, z)$ to be a $\Gamma \to \Delta$ partial injection, independent
of $\texttt{p}(X, Y, z')$ for any $z' \ne z$. \textsc{Crane} instantly returns a
solution of complexity $\Theta(l+mn)$, where $l = |\Lambda|$.

@@

In this paper, we showed how a state-of-the-art (W)FOMC algorithm can be
empowered by generalising domain recursion and adding support for cycles in the
graph that encodes a solution. To construct such graphs, \textsc{Crane}
supplements \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} with three new
compilation rules. Our results in \cref{sec:results} revealed a range of
counting problems that are liftable to \textsc{Crane} but not to any other
(W)FOMC algorithm. Although in this paper we focus on unweighted counting,
\textsc{ForcLift}'s support for weights trivially transfers to \textsc{Crane} as
well. The common thread across these newly liftable problems is (partial)
injectivity. Thus, we can formulate the following conjecture.

@@

% TODO: argue about the possibilities of optimising the recurrence formula
% further, relate this to skolem function synthesis (maybe elsewhere as well)
